% ##TransversalGates
\documentclass[a4paper, english]{scrartcl}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{babel}
\usepackage[cm]{fullpage}
\usepackage{float}
\usepackage{graphicx}
\usepackage{helvet}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{nicefrac}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{placeins}
\usepackage{verbatim}
\usepackage{xcolor}
\definecolor{gr}{gray}{0.9}
\renewcommand{\familydefault}{\sfdefault}
\title{Quantum Error Correction}
\subtitle{Transversal Cliffords on $\qec{n}{n-2}{2}$ Codes}
\author{Ben Criger}
\date{\today}
\input{../../../Qcircuit.tex}
\input{bencommands.tex}
\newlength\figureheight
\newlength\figurewidth
\setlength\figureheight{7cm}
\setlength\figurewidth{12cm}

\providecommand{\cnot}{\textsc{cnot}}
\providecommand{\chieff}{\chi_{\textrm{eff}}}

\usetikzlibrary{decorations.pathreplacing,decorations.pathmorphing}

\begin{document}
\maketitle
\section{Introduction}
It has been noted that there exists an $\qec{8}{3}{2}$ code in which a transversal $T (= R_{z,\nicefrac{\pi}{8}})$ gate produces a doubly-controlled $Z$ on the logical qubits. 
It would be nice to generalize this result to non-cube structures, or even larger square tilings of the sphere. 
To begin, I'm going to recap the equivalent result in 2D, then generalize it to non-squares. 

\section{Transversal Phase Gate in the $\qec{4}{2}{2}$ Code}
The stabilizers and logical operators of the $\qec{4}{2}{2}$ code should be old hat by now:
\begin{equation}
S = \group{S_X = XXXX, S_Z = ZZZZ}, \quad L = \group{\overline{X}_0 = XXII,\, \overline{X}_1 = XIXI,\, \overline{Z}_0 = IZIZ,\, \overline{Z}_1 = IIZZ}
\end{equation}
The action of a bipartite phase gate $P \otimes P\ct \otimes P\ct \otimes P $ (where $P = R_{z,\nicefrac{\pi}{4}}$) preserves the stabilizer, since $XXXX$ can be regenerated by multiplying $YYYY$ and $ZZZZ$. 
The effect on the logical $X$ operators is to map them to $-YYII$ and $-YIYI$.
Multiplying these new operators by $S_Z$ takes them to $XXZZ = \overline{X}_0 \overline{Z}_1$ and $XZXZ = \overline{X}_1 \overline{Z}_0$, performing a \textsc{cphase} gate on the logical qubits. 
In the following section, we generalize this to codes on even-weight polygons.

\section{Polygonal Phase Gate}
We begin again with two stabilizers, $S_Z = Z^{\otimes n}$  and $S_X = X^{\otimes n}$, producing an $\qec{n}{n-2}{2}$ code. 
If $n=0\mod 4$, we can perform a similar phase gate $P \otimes P^{\dagger \otimes n - 2} \otimes P $, and the stabilizer will be preserved, since it is preserved under $P^{\otimes 2} \otimes P^{\dagger \otimes 2}$ and $P^{\dagger \otimes 4}$ separately. 
If $n=2\mod 4$, and we applied the same gate, we would end up with $-X^{\otimes n}$ in the stabilizer.
In this case, we apply the gate $P \otimes P^{\dagger \otimes n - 1}$ instead. 

The $n-2$ pairs of logical operators can be written as $\overline{X}_j = X_0 X_j$, $\overline{Z}_j = Z_j Z_{n-1}$ for $j$ between $1$ and $n-2$. 
As before, $X_0 X_j \mapsto -Y_0 Y_j$. 
If we multiply by $S_Z$, the result is $X_0 X_j \cdot \bigotimes_{k \neq 0, j} Z_k$.
It is important to note that the set $\set{k \suchthat k\neq 0,j}$ is even-size, so that the set $\set{k \suchthat k\neq 0,j,n-1}$ is odd-size. 
This implies that, if we take the product of $\overline{Z_k}, k \neq j$, the qubit $n-1$ will appear in the argument an odd number of times, and therefore be in the support of the product. 
This implies that the transversal gate described above takes $\overline{X}_j$ to $\overline{X}_j \cdot \bigotimes_{k \neq j} \overline{Z}_j$, performing a complete graph's worth of \textsc{cphase} gates in a single timestep. 

\section{Future Work}
\begin{itemize}
\item Hadamards
\item Polyhedra
\item Replace each face of the cube with four squares. 
\end{itemize}

\end{document} 